Wait-free Queues with Polygarithmic Step Complexity

### 2022-09-29 backup/jul12.tex
fix citations, review index deq

\algdef{S}[FUNCTION]{Function}
   [3]{{\texttt{\textsl{#1}}} \textproc{\texttt{#2}}\ifthenelse{\equal{#3}{}}{}{\texttt{(#3)}}}

@@

\Statex $\blacktriangleright$ \texttt{\textsl{operation} op}
\begin{itemize}
%\item \texttt{\textsl{string} name}
\item \texttt{\textsl{int} pid}
\item \texttt{\textsl{int} loc\textsf{: location in \texttt{l\textsubscript{pid}}'s array}}
\end{itemize}

\Statex $\blacktriangleright$ \texttt{\textsl{leaf} l\textsubscript{i}}
\begin{itemize}
\item \texttt{\textsl{*node} parent}\textsf{: pointer to parent in Tree}
\item \texttt{\textsl{operation[]} ops}\textsf{: ops invoked by process i}
\item \texttt{\textsl{int} last= 1}\textsf{: index of first empty cell of \texttt{ops}}
\end{itemize}

\Statex $\blacktriangleright$ \texttt{\textsl{node} n}
\begin{itemize}
\item \texttt{\textsl{*node} left, right, parent}
\item \texttt{\textsl{block[]} blocks}\textsf{: blocks propagated to \texttt{n}}
\item \texttt{\textsl{int} last= 1}\textsf{: index of first empty cell of \texttt{blocks}}
\item \texttt{\textsl{int} counter= 0}\textsf{: \#groups of blocks propagated from \texttt{n}}
\item \texttt{\textsl{int[]} super}\textsf{: approximate index of superblock of a block in the given group in \texttt{n}'s parent}
\end{itemize}

\Statex $\blacktriangleright$ \texttt{\textsl{block} b}
\begin{itemize}
  \item \texttt{\textsl{int} num\textsubscript{left}, num\textsubscript{right}, sum, sum\textsubscript{left}, sum\textsubscript{right}}\textsf{: \#operations from the left and the right child, sum of all operations, prefix sum of operations from the left and the right child}
  \item \texttt{\textsl{int} end\textsubscript{left}, end\textsubscript{right}}\textsf{: index of \texttt{b}'s last subblock}
  \item \texttt{\textsl{int} group}\textsf{: approximate group number}
\end{itemize}

\Statex \textbf{Examples}
\Statex \hspace{0pt} \texttt{ l\textsubscript{i}[j]}\textsf{: jth operation of ith process}
\Statex \hspace{0pt} \texttt{ n.left}\textsf{: left child of node n}
\Statex \hspace{0pt} \texttt{ n.blocks[n.last]}\textsf{: index of last block of node n}
\Statex \hspace{0pt} \texttt{ n.blocks[i].num\textsubscript{left}}\textsf{: \#left operations of ith block of n}
\Statex \hspace{4pt} \texttt{b.end\textsubscript{right}}\textsf{: index of ending right subblock of b}

@@

\Function{int}{Append}{\textsl{operation} op, \textsl{int} pid} 
\State \texttt{op.loc= l\textsubscript{pid}.last + 1}
\State \texttt{l\textsubscript{pid}.ops[op.location]= op}
\State \Call{Propagate}{l\textsubscript{pid}.parent}
\State \Return \Call{Index}{l\textsubscript{pid}, op.location, b} \Comment we don't know the block eachoperation resides in leaf
\EndFunction{Append}

@@

\Function{void}{Propagate}{\textsl{node} n}
\If{\texttt{n {\keywordfont is} root}} \Return
\ElsIf{\texttt{not} \Call{Refresh}{n}}
\State \Call{Refresh}{n}
\EndIf
\State \Call{Propagate}{n.parent}
\EndFunction{Propagate}

@@

\Function{boolean}{Refresh}{\textsl{node} n}
\State \texttt{i= n.last}
\State \texttt{new, c\textsubscript{left}, c\textsubscript{right}= \Call{CreateBlock}{n, i+1, n.counter}}
\If{\Call{CAS}{n.blocks[i+1], null, new}}
\State \texttt{\Call{CAS}{n.last, i, i+1}}
\ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}}
\State \texttt{\Call{CAS}{n.dir.super[c\textsubscript{dir}], null, i+1}}
\State \texttt{\Call{CAS}{n.dir.counter, c\textsubscript{dir}, c\textsubscript{dir}+1}}
\EndFor
\State \Return{\texttt{true}}
\Else ~\Return{ \texttt{false}}
\EndIf
\EndFunction{Index}

@@

\Function{block, int, int}{CreateBlock}{\textsl{node} n, \textsl{int} i, int g} 
\Statex\Comment{Creates a block to insert into \texttt{n.blocks[i]}}
\State \texttt{block b= \Call{new}{\textsl{block}}}
\If{\texttt{n is leaf}}\Comment{TODO}
\Else
\State \texttt{b.group= g}
\ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}}
\State \texttt{j= n.dir.last}
\State \texttt{start= n.blocks[i-1].end\textsubscript{dir}}
\State \texttt{end= n.dir.blocks[j]}
\State \texttt{c\textsubscript{dir}= n.dir.blocks[j].group}
\State \texttt{b.end\textsubscript{dir}= j}
\State \texttt{b.num\textsubscript{dir}= end.sum\textsubscript{\#dir} - start.sum\textsubscript{\#dir}}
\State \texttt{b.sum\textsubscript{\#dir}= n.blocks[i-1].sum\textsubscript{\#dir} + b.num\textsubscript{dir}}
\EndFor
\State \texttt{b.sum= b.sum\textsubscript{left} + b.sum\textsubscript{right}}
\EndIf
\State \Return \texttt{b, c\textsubscript{left}, c\textsubscript{rightt}}
\EndFunction{CreateBlock}

@@

\Function{element}{Get}{\textsl{int} i} 
\State \texttt{res= \Call{BSearch}{root, sum, i, 0, root.last}}
\State \Return{\Call{Get}{root, res, i-root.blocks[res-1].sum}}
\EndFunction{Get}

@@

\Function{element}{Get}{\textsl{node} n, \textsl{int} b, \textsl{int} i} 
\Comment{\textmd{Returns the \texttt{i}th operation in \texttt{b}th block of node \texttt{n}}}

\If{\texttt{i $\leq$ n.blocks[b].num\textsubscript{left}}} \Comment{\texttt{i} exists in left child of \texttt{n}}
\If{\texttt{n.left {\keywordfont is} leaf}} \Return \texttt{n.left.ops[i]}
\Else
\State \texttt{subBlock= \Call{BSearch}{n.left.sum, i, n.blocks[b-1].end\textsubscript{left}+1, n.blocks[b].end\textsubscript{left}}}
\State \Return\Call{Get}{n.left, subBlock, i-n.left.blocks[subBlock-1].sum} 
\EndIf
\Else
\State \texttt{i= i-n.blocks[b].num\textsubscript{left}}
\If{\texttt{n.right {\keywordfont is} leaf}} \Return \texttt{n.right.ops[i]}
\Else 
\State\texttt{subBlock=\Call{BSearch}{n.right.sum, i, n.blocks[b-1].end\textsubscript{right}, n.blocks[b].end\textsubscript{right}}}
\State \Return\Call{Get}{n.right, subBlock, i-n.right.blocks[subBlock-1].sum} \EndIf
\EndIf
\EndFunction{Get}

@@

\Function{index}{Index}{\textsl{node} n, \textsl{int} i, \textsl{int} b} \Comment{Returns order in  the root of \texttt{i}th operation in the node \texttt{n} which exists in block b}
\If{\texttt{n {\keywordfont is} root}} \Return \texttt{i}
\Else
\State \texttt{dir= (n.parent.left==n)? left: right}
\State \texttt{superBlock= \Call{BSearch}{n.parent, n.sum\textsubscript{dir}, i, sup[n.blocks[b].group]-p, sup[n.blocks[b].group]+p}}
\If{\texttt{dir {\keywordfont is} left}}
\State \texttt{i+= n.parent.blocks[superBlock-1].sum\textsubscript{right}}
\Else \State \texttt{i+= n.parent.blocks[superBlock].sum\textsubscript{left} + n.parent.blocks[superBlock].sum\textsubscript{left} + n.blocks[nBlock-1].sum}
\EndIf
\State \Return\Call{Index}{n.parent, i, superBlock}
\EndIf
\EndFunction{Index}

>>>>

\algdef{S}[FUNCTION]{Function}
   [3]{{\keywordfont #1} \textproc{\texttt{#2}}\ifthenelse{\equal{#3}{}}{}{\texttt{(#3)}}}

@@

\Statex $\blacktriangleright$ \texttt{\textsl{operation} op}
\begin{itemize}
\item \texttt{\textsl{string} name}
\item \texttt{\textsl{int} pid}
\item \texttt{\textsl{int} loc \textsf{: location in leaf's array}}
\end{itemize}

\Statex $\blacktriangleright$ \texttt{\textsl{leaf} l\textsubscript{i}}
\begin{itemize}
\item \texttt{\textsl{*node} parent} \textsf{: pointer to parent in Tree}
\item \texttt{\textsl{operation[]} ops} \textsf{: ops invoked by process i}
\item \texttt{\textsl{int} last= 1} \textsf{: index of first empty cell of \texttt{ops}}
\end{itemize}

\Statex $\blacktriangleright$ \texttt{\textsl{node} n}
\begin{itemize}
\item \texttt{\textsl{*node} left, right, parent}
\item \texttt{\textsl{block[]} blocks} \textsf{: blocks propagated to \texttt{n}}
\item \texttt{\textsl{int} last= 1} \textsf{: index of first empty cell of \texttt{blocks}}
\end{itemize}

\Statex $\blacktriangleright$ \texttt{\textsl{block} b}
\begin{itemize}
  \item \texttt{\textsl{int} num\textsubscript{left}, num\textsubscript{right}, sum, sum\textsubscript{left}, sum\textsubscript{right}} \textsf{:~\#~operations from the left and the right child, sum of all operations, prefix sum of operations from the left and the right child}
  \item \texttt{\textsl{int} end\textsubscript{left}, end\textsubscript{right}} \textsf{: index of \texttt{b}'s last subblock}
\end{itemize}

\Statex \textbf{Examples}
\Statex \hspace{0pt} \texttt{ l\textsubscript{i}[j]}\textsf{: jth operation of ith process}
\Statex \hspace{0pt} \texttt{ n.left}\textsf{: left child of node n}
\Statex \hspace{0pt} \texttt{ n.blocks[n.last]}\textsf{: index of last block of node n}
\Statex \hspace{0pt} \texttt{ n.blocks[i].num\textsubscript{left}}\textsf{: \#left operations of ith block of n}
\Statex \hspace{4pt} \texttt{b.end\textsubscript{right}}\textsf{: ending right subblock of b position}

@@

\Function{int}{Append}{\textsl{operation} op, \textsl{int} pid} 
\State \texttt{op.loc= l\textsubscript{pid}.last + 1}
\State \texttt{l\textsubscript{pid}.ops[op.location]= op}
\State \Call{Propagate}{l\textsubscript{pid}.parent}
\State \Return \Call{Index}{l\textsubscript{pid}, op.location}
\EndFunction{Append}

@@

\Function{element}{Get}{\textsl{int} i} 
\State \texttt{res=\Call{BSearch}{root, sum, i, 0, root.last}}
\State \Return{\Call{Get}{root, res, i-root.blocks[res-1].sum}}
\EndFunction{Get}

@@

%\Function{index}{Index}{operation op, int pid}
%\State \texttt{\textsl{int} i= op's order in l\textsubscript{pid}.ops}
%\State \Return \Call{Index}{l\textsubscript{p}, i}
%\EndFunction{Index}

%\Statex

\Function{void}{Propagate}{\textsl{node} n}
\If{\texttt{n {\keywordfont is} root}} \Return
\Else
\State \texttt{i= n.last}
\State \texttt{\textsl{block} new=\Call{CreateBlock}{n, i+1}}
\If{\Call{CAS}{n.blocks[i+1], null, new}}
\State \texttt{\Call{CAS}{n.last, i, i+1}}
\Else
\State \texttt{i= n.last}
\State \texttt{\textsl{block} new=\Call{CreateBlock}{n, i+1}}
\If{\Call{CAS}{n.blocks[i+1], null, new}}
\State \texttt{\Call{CAS}{n.last, i, i+1}}
\EndIf
\EndIf
\EndIf
\State \Call{Propagate}{n.parent}
\EndFunction{Propagate}

@@

\Function{block}{CreateBlock}{\textsl{node} n, \textsl{int} i} 
\Statex\Comment{Creates a block to insert into \texttt{n.blocks[i]}}
\State \texttt{block b= \Call{new}{\textsl{block}}}
\ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}}
\State \texttt{prev= n.blocks[i-1].ending\textsubscript{dir}}
\State \texttt{j= n.dir.last}
\State \texttt{end= n.dir.blocks[j]}
\State \texttt{b.end\textsubscript{dir}= j}
\State \texttt{b.num\textsubscript{dir}= end.sum\textsubscript{\#dir} - start.sum\textsubscript{\#dir}}
\State \texttt{b.sum\textsubscript{\#dir}= n.blocks[i-1].sum\textsubscript{\#dir} + b.num\textsubscript{dir}}
\EndFor
\State \texttt{b.sum= b.sum\textsubscript{left} + b.sum\textsubscript{right}}
\State \Return \texttt{b}
\EndFunction{CreateBlock}

@@

@@

\Function{element}{Get}{node n, int b, int i} 
\Comment{\textmd{Returns the \texttt{i}th operation in \texttt{b}th block of node \texttt{n}}}

\If{\texttt{i $\leq$ b.num\textsubscript{left}+1}} \Comment{\texttt{i} exists in left child of \texttt{n}}
\If{\texttt{n.left {\keywordfont is} leaf}} \Return \texttt{n.left.ops[i]}
\Else
\State \texttt{subBlock= \Call{BSearch}{n.left.sum, i, n.blocks[b-1].end\textsubscript{left}+1, n.blocks[b].end\textsubscript{left}}}
\State \Return\Call{Get}{n.left, subBlock, i-n.left.blocks[subBlock-1].sum} 
\EndIf
\Else
\State \texttt{i= i-n.blocks[b].sum\textsubscript{left}}
\If{\texttt{n.right {\keywordfont is} leaf}} \Return \texttt{n.right.ops[i]}
\Else 
\State\texttt{subBlock=\Call{BSearch}{n.right.sum, i, n.blocks[b-1].ending\textsubscript{right}, n.blocks[b].ending\textsubscript{right}}}
\State \Return\Call{Get}{n.right, subBlock, i-n.right.blocks[subBlock-1].sum} \EndIf
\EndIf
\EndFunction{Get}

@@

\Function{index}{Index}{node n, int i} \Comment{Returns order in root of \texttt{i}th operation in node \texttt{n}}
\If{\texttt{n {\keywordfont is} root}} \Return \texttt{i}
\Else
\State \texttt{nBlock= \Call{BSearch}{n, n.sum, i, 0, n.parent.last}}
\Comment{\textmd{This search takes O(log n) steps}}
\State \texttt{dir= (n.parent.left==n)?left:right}
\State \texttt{superBlock= \Call{BSearch}{n.parent, n.sum\textsubscript{dir}, i, 0, n.parent.last}}
\Comment{\textmd{This search takes O(log n) steps}}
\If{\texttt{dir {\keywordfont is} left}}
\State \texttt{i+= n.parent.blocks[superBlock-1].sum + n.blocks[nBlock-1].sum}
\Else \State \texttt{i+= n.parent.blocks[superBlock-1].sum + n.parent.blocks[superBlock].sum\textsubscript{left} + n.blocks[nBlock-1].sum}
\EndIf
\State \Return\Call{Index}{n.parent,}
\EndIf
\EndFunction{Index}
